---
title: "Statistical analyses: lignocellulosic inhibitors and lactic acid bacteria contamination in *S. cerevisiae* PE-2 co-cultures"
author: "Thamiris Guerra Giacon"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

## Install packages

```{r}
install.packages(c("DescTools", "tidyr")) 
```

```{r}
library(DescTools)
library(tidyr)
library(tidyr)
library(dplyr)
library(multcomp)
library(ggplot2)y
```

\# Overview This document contains the statistical analyses supporting the manuscript "[manuscript title]". It is intended to make every test reported in the paper reproducible from the raw data.

## Experimental design

The industrial yeast *Saccharomyces cerevisiae* PE-2 was cultivated in MBL medium (20 g/L glucose, pH 6.0, 30 °C, static, 24 h) under four inoculation conditions:

|           |                     |                    |                    |
|-----------|---------------------|--------------------|--------------------|
| **Assay** | ***S. cerevisiae*** | ***L. fermentum*** | ***L. plantarum*** |
| **A**     | 10⁷                 | –                  | –                  |
| **B**     | 10⁷                 | 10⁷                | –                  |
| **C**     | 10⁷                 | –                  | 10⁷                |
| **D**     | 10⁷                 | 5 × 10⁶            | 5 × 10⁶            |

Each inoculation condition was combined with one of ten inhibitor conditions: no inhibitor (control), HMF, furfural, acetic, formic, glycolic, levulinic and ferulic acids, syringaldehyde and vanillin, each tested individually. \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

# Fig 1. Yeast viability

Yeast viability (%) was determined after 24 h by methylene blue staining and counting in a Neubauer chamber. The question addressed here is whether bacterial contamination reduces yeast viability relative to the bacteria-free culture exposed to the same inhibitor, and whether the two bacterial strains differ in their effect.

```{r}

## ---- viability-data ----
wide <- read.csv("viability.csv", check.names = FALSE)
 
inhibitor_levels <- c("Control", "HMF", "Furfural", "Acetic acid", "Formic acid",
                      "Glycolic acid", "Levulinic acid", "Ferulic acid",
                      "Syringaldehyde", "Vanillin")
assay_levels <- c("none", "LF", "LP", "LFLP")
 
viab <- wide %>%
  pivot_longer(-inhibitor, names_to = c("assay", "rep"),
               names_sep = "_", values_to = "viability") %>%
  mutate(inhibitor = factor(inhibitor, levels = inhibitor_levels),
         assay     = factor(assay, levels = assay_levels))

## ---- viability-stats ----
# Within each inhibitor block: one-way ANOVA, Dunnett vs bacteria-free (none),
# Tukey among contaminated assays (LF, LP, LF+LP).
anova_res <- list(); dunnett_res <- list(); tukey_res <- list()
 
for (inh in inhibitor_levels) {
  d <- filter(viab, inhibitor == inh)
  fit <- aov(viability ~ assay, data = d)
  anova_res[[inh]] <- data.frame(inhibitor = inh,
                                 F = summary(fit)[[1]][["F value"]][1],
                                 p = summary(fit)[[1]][["Pr(>F)"]][1])
 
  dun <- summary(glht(fit, linfct = mcp(assay = "Dunnett")))
  dunnett_res[[inh]] <- data.frame(inhibitor  = inh,
                                   comparison = names(dun$test$coefficients),
                                   diff       = unname(dun$test$coefficients),
                                   p_adj      = unname(dun$test$pvalues))
 
  dc  <- droplevels(filter(d, assay != "none"))
  tuk <- TukeyHSD(aov(viability ~ assay, data = dc))$assay
  tukey_res[[inh]] <- data.frame(inhibitor = inh, comparison = rownames(tuk),
                                 diff = tuk[, "diff"], p_adj = tuk[, "p adj"])
}
 
anova_tab   <- bind_rows(anova_res)
dunnett_tab <- bind_rows(dunnett_res)
tukey_tab   <- bind_rows(tukey_res); rownames(tukey_tab) <- NULL
 
knitr::kable(anova_tab,   digits = 3, caption = "One-way ANOVA per inhibitor")
knitr::kable(dunnett_tab, digits = 3, caption = "Dunnett's test vs bacteria-free control")
knitr::kable(tukey_tab,   digits = 3, caption = "Tukey's HSD among contaminated assays")
 
## ---- viability-figure ----
# Summary (mean, SD) and significance marks from Dunnett's test
sig <- dunnett_tab %>%
  mutate(assay = sub(" - none", "", comparison),
         star  = ifelse(p_adj < 0.05, "*", "")) %>%
  dplyr::select(inhibitor, assay, star)
 
summ <- viab %>%
  group_by(inhibitor, assay) %>%
  summarise(mean = mean(viability), sd = sd(viability), .groups = "drop") %>%
  left_join(mutate(sig, inhibitor = factor(inhibitor, levels = inhibitor_levels),
                   assay = factor(assay, levels = assay_levels)),
            by = c("inhibitor", "assay")) %>%
  mutate(star       = ifelse(is.na(star), "", star),
         label_mean = paste0(sprintf("%.0f", mean), star),
         label_sd   = paste0("± ", sprintf("%.1f", sd)),
         text_col   = ifelse(mean > 88, "white", "grey10"))
 
assay_labels <- c(none = "No LAB", LF = "LF", LP = "LP", LFLP = "LF + LP")
x_labels <- c("Control",
              expression("HMF (1 g·L"^-1*")"),
              expression("Furfural (3 g·L"^-1*")"),
              expression("Acetic acid (4 g·L"^-1*")"),
              expression("Formic acid (0.75 g·L"^-1*")"),
              expression("Glycolic acid (0.75 g·L"^-1*")"),
              expression("Levulinic acid (3 g·L"^-1*")"),
              expression("Ferulic acid (0.75 g·L"^-1*")"),
              expression("Syringaldehyde (0.75 g·L"^-1*")"),
              expression("Vanillin (0.75 g·L"^-1*")"))
 
fig1 <- ggplot(summ, aes(x = inhibitor, y = assay, fill = mean)) +
  geom_tile(colour = "white", linewidth = 1.2) +
  geom_text(aes(label = label_mean, colour = text_col),
            nudge_y = 0.13, size = 3.4, fontface = "bold") +
  geom_text(aes(label = label_sd, colour = text_col),
            nudge_y = -0.20, size = 2.5) +
  scale_colour_identity() +
  scale_fill_distiller(palette = "Blues", direction = 1, limits = c(65, 100),
                       name = "Yeast viability (%)") +
  scale_y_discrete(limits = rev(assay_levels), labels = assay_labels) +
  scale_x_discrete(labels = x_labels) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 10) +
  theme(panel.grid  = element_blank(),
        axis.text.x = element_text(angle = 40, hjust = 1, colour = "black"),
        axis.text.y = element_text(colour = "black"),
        legend.title = element_text(angle = 90, hjust = 0.5),
        legend.key.height = unit(1.2, "cm")) +
  guides(fill = guide_colourbar(title.position = "right"))
 
fig1
 
ggsave("Fig1_viability.tiff", fig1, width = 22, height = 9, units = "cm",
       dpi = 600, device = "tiff", compression = "lzw", bg = "white")
ggsave("Fig1_viability.pdf", fig1, width = 22, height = 9, units = "cm",
       device = cairo_pdf, bg = "white")
 
```

\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

# Fig2. yeast growth

```{r}
## ---- growth-setup ----
library(readxl)
library(tidyr)
library(dplyr)
library(multcomp)
library(ggplot2)
 
## ---- growth-data ----
# Same input file used for the bubble plot: one row per inhibitor,
# yeast growth ratio (X24/X0) for each assay and replicate.
df_raw <- read_excel(
  path      = "bubbleplot.xlsx",
  sheet     = 1,
  skip      = 2,
  col_names = c("inhibitor",
                "none_1", "none_2", "none_3",
                "LF_1",   "LF_2",   "LF_3",
                "LP_1",   "LP_2",   "LP_3",
                "LFLP_1", "LFLP_2", "LFLP_3")
) %>%
  filter(!is.na(inhibitor), inhibitor != "") %>%
  mutate(across(-inhibitor, as.numeric))
 
inhibitor_levels <- df_raw$inhibitor          # keeps the order of the spreadsheet
assay_levels     <- c("none", "LF", "LP", "LFLP")
 
growth <- df_raw %>%
  pivot_longer(-inhibitor, names_to = c("assay", "rep"),
               names_sep = "_", values_to = "ratio") %>%
  mutate(inhibitor = factor(inhibitor, levels = inhibitor_levels),
         assay     = factor(assay, levels = assay_levels),
         rep       = factor(rep))
 
## ---- growth-normalization ----
# Normalization to the bacteria-free culture of the same block.
# "paired": replicate i of each assay divided by replicate i of No LAB
#           (use only if replicate i of all assays came from the same yeast inoculum)
# "mean":   each replicate divided by the mean of No LAB in the same block
normalization <- "mean"
 
growth <- growth %>%
  group_by(inhibitor, rep) %>%
  mutate(ref_paired = ratio[assay == "none"]) %>%
  group_by(inhibitor) %>%
  mutate(ref_mean = mean(ratio[assay == "none"])) %>%
  ungroup() %>%
  mutate(fc = if (normalization == "paired") ratio / ref_paired else ratio / ref_mean)
 
## ---- growth-stats ----
# All replicates of a block were inoculated from a single pre-culture, so
# replicate numbering is arbitrary: no pairing, no replicate factor.
# Tests are run on log-transformed raw growth ratios, log(X24/X0), not on
# normalized values: the normalized No LAB group would have no variance, and
# the log removes the dependence of the variance on the mean (ratios are multiplicative).
anova_res <- list(); dunnett_res <- list(); tukey_res <- list()
 
for (inh in inhibitor_levels) {
  d   <- filter(growth, inhibitor == inh)
  fit <- aov(log(ratio) ~ assay, data = d)
  anova_res[[inh]] <- data.frame(inhibitor = inh,
                                 F = summary(fit)[[1]][["F value"]][1],
                                 p = summary(fit)[[1]][["Pr(>F)"]][1])
 
  dun <- summary(glht(fit, linfct = mcp(assay = "Dunnett")))
  dunnett_res[[inh]] <- data.frame(inhibitor  = inh,
                                   comparison = names(dun$test$coefficients),
                                   diff       = unname(dun$test$coefficients),
                                   p_adj      = unname(dun$test$pvalues))
 
  dc  <- droplevels(filter(d, assay != "none"))
  tuk <- TukeyHSD(aov(log(ratio) ~ assay, data = dc))$assay
  tukey_res[[inh]] <- data.frame(inhibitor = inh, comparison = rownames(tuk),
                                 diff = tuk[, "diff"], p_adj = tuk[, "p adj"])
}
 
growth_anova   <- bind_rows(anova_res)
growth_dunnett <- bind_rows(dunnett_res)
growth_tukey   <- bind_rows(tukey_res); rownames(growth_tukey) <- NULL
 
knitr::kable(growth_anova,   digits = 3, caption = "One-way ANOVA per inhibitor, log(X24/X0)")
knitr::kable(growth_dunnett, digits = 3, caption = "Dunnett's test vs bacteria-free control")
knitr::kable(growth_tukey,   digits = 3, caption = "Tukey's HSD among contaminated assays")
 
## ---- growth-figure ----
sig <- growth_dunnett %>%
  mutate(assay = sub(" - none", "", comparison),
         star  = ifelse(p_adj < 0.05, "*", "")) %>%
  dplyr::select(inhibitor, assay, star) %>%
  mutate(inhibitor = factor(inhibitor, levels = inhibitor_levels),
         assay     = factor(assay, levels = assay_levels))
 
summ <- growth %>%
  group_by(inhibitor, assay) %>%
  summarise(mean = mean(fc), sd = sd(fc), .groups = "drop") %>%
  left_join(sig, by = c("inhibitor", "assay")) %>%
  mutate(star       = ifelse(is.na(star), "", star),
         label_mean = paste0(sprintf("%.2f", mean), star),
         label_sd   = ifelse(assay == "none" & normalization == "paired", "",
                             paste0("± ", sprintf("%.2f", sd))),
         log2fc     = log2(mean),
         )
 
# Symmetric color limits in log2 space, centered on fold-change = 1
lim <- max(abs(summ$log2fc), na.rm = TRUE)
summ <- mutate(summ, text_col = ifelse(abs(log2fc) > 0.6 * lim, "white", "grey10"))
fc_breaks <- c(0.25, 0.5, 1, 2, 4)
fc_breaks <- fc_breaks[abs(log2(fc_breaks)) <= lim]
 
assay_labels <- c(none = "No LAB", LF = "LF", LP = "LP", LFLP = "LF + LP")
 
# Bubble plot: bubble size and color = mean normalized growth;
# label inside the bubble = mean (+ * if significant); SD below the bubble.
fig2 <- ggplot(summ, aes(x = inhibitor, y = assay)) +
  geom_point(aes(size = mean, fill = log2fc),
             shape = 21, colour = "grey30", stroke = 0.4) +
  geom_text(aes(label = label_mean, colour = text_col),
            size = 2.7, fontface = "bold") +
  geom_text(aes(label = label_sd), colour = "grey35",
            nudge_y = -0.42, size = 2.3) +
  scale_colour_identity() +
  scale_size_continuous(range = c(11, 19), guide = "none") +  # min size keeps labels inside
  scale_fill_distiller(palette = "RdBu", direction = 1, limits = c(-lim, lim),
                       breaks = log2(fc_breaks), labels = fc_breaks,
                       name = "Yeast growth\n(relative to No LAB)") +
  scale_y_discrete(limits = rev(assay_levels), labels = assay_labels,
                   expand = expansion(add = 0.6)) +
  scale_x_discrete(guide = guide_axis(angle = 40)) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 10) +
  theme(panel.grid.major = element_line(colour = "grey92", linewidth = 0.3),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "grey85", fill = NA, linewidth = 0.4),
        axis.text   = element_text(colour = "black"),
        legend.title = element_text(size = 8.5),
        plot.margin = margin(12, 8, 4, 4),
        legend.key.height = unit(1.2, "cm"))
 
fig2
 
ggsave("Fig2_growth.tiff", fig2, width = 22, height = 11, units = "cm",
       dpi = 600, device = "tiff", compression = "lzw", bg = "white")
ggsave("Fig2_growth.pdf", fig2, width = 22, height = 11, units = "cm",
       device = cairo_pdf, bg = "white")
```

\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

# Fig3. yeast viability vs growth

```{r}
## ---- vg-setup ----
library(readxl)
library(tidyr)
library(dplyr)
library(multcomp)
library(ggplot2)
library(ggrepel)
 
## ---- vg-data ----
# Inputs: viability.csv (yeast viability, %) and bubbleplot.xlsx (yeast X24/X0).
# Both files list the ten inhibitor conditions in the same order (control first).
inh_lab   <- c("Control", "HMF", "Furfural", "Acetic acid", "Formic acid",
               "Glycolic acid", "Levulinic acid", "Ferulic acid",
               "Syringaldehyde", "Vanillin")
inh_class <- c("Control", "Furan", "Furan", "Weak acid", "Weak acid", "Weak acid",
               "Weak acid", "Phenolic", "Phenolic", "Phenolic")
 
viab <- read.csv("viability.csv", check.names = FALSE) %>%
  mutate(inhibitor = inh_lab) %>%
  pivot_longer(-inhibitor, names_to = c("assay", "rep"), names_sep = "_", values_to = "y") %>%
  mutate(assay = factor(assay, levels = c("none", "LF", "LP", "LFLP")))
 
grow <- read_excel("bubbleplot.xlsx", skip = 2,
                   col_names = c("inhibitor", paste0(rep(c("none", "LF", "LP", "LFLP"), each = 3),
                                                     "_", 1:3))) %>%
  filter(!is.na(inhibitor), inhibitor != "") %>%
  mutate(inhibitor = inh_lab, across(-inhibitor, as.numeric)) %>%
  pivot_longer(-inhibitor, names_to = c("assay", "rep"), names_sep = "_", values_to = "y") %>%
  mutate(assay = factor(assay, levels = c("none", "LF", "LP", "LFLP")))
 
## ---- vg-summary ----
# x: change in viability vs No LAB of the same block (percentage points)
# y: yeast growth (X24/X0) relative to the mean of No LAB of the same block
# sig: Dunnett's test vs No LAB within each block (viability on %, growth on log ratio)
dunnett_p <- function(d, formula) {
  p <- summary(glht(aov(formula, data = d), linfct = mcp(assay = "Dunnett")))$test$pvalues
  data.frame(assay = c("LF", "LP", "LFLP"), p = as.numeric(p))
}
 
vg <- lapply(inh_lab, function(i) {
  dv <- filter(viab, inhibitor == i); dg <- filter(grow, inhibitor == i)
  ref_v <- mean(dv$y[dv$assay == "none"]); ref_g <- mean(dg$y[dg$assay == "none"])
  sv <- dv %>% filter(assay != "none") %>% group_by(assay) %>%
    summarise(dviab = mean(y) - ref_v, .groups = "drop") %>%
    left_join(rename(dunnett_p(dv, y ~ assay), p_viab = p), by = "assay")
  sg <- dg %>% filter(assay != "none") %>% group_by(assay) %>%
    summarise(growth = mean(y) / ref_g, .groups = "drop") %>%
    left_join(rename(dunnett_p(dg, log(y) ~ assay), p_growth = p), by = "assay")
  left_join(sv, sg, by = "assay") %>% mutate(inhibitor = i)
}) %>% bind_rows() %>%
  mutate(class = factor(inh_class[match(inhibitor, inh_lab)],
                        levels = c("Control", "Furan", "Weak acid", "Phenolic")),
         assay = factor(recode(as.character(assay), LFLP = "LF + LP"),
                        levels = c("LF", "LP", "LF + LP")),
         sig_growth = factor(ifelse(p_growth < 0.05, "p < 0.05", "n.s."), levels = c("p < 0.05", "n.s.")),
         label = ifelse(p_viab < 0.05, paste0(inhibitor, " \u2020"), inhibitor))
 
knitr::kable(vg, digits = 3, caption = "Viability change and growth ratio vs No LAB")
 
## ---- vg-figure ----
strain_cols <- c("LF" = "#F9BC3C",        # L. fermentum (yellow)
                 "LP" = "#F8866D",        # L. plantarum (salmon)
                 "LF + LP" = "#7A5340")   # both strains (brown)
class_shapes <- c("Control" = 22, "Furan" = 21, "Weak acid" = 24, "Phenolic" = 23)
# Thick black outline = significant effect on growth;
# dagger (\u2020) after the label = significant effect on viability
sig_stroke   <- c("p < 0.05" = 1.3, "n.s." = 0.35)
sig_col      <- c("p < 0.05" = "black", "n.s." = "grey45")
 
fig_vg <- ggplot(vg, aes(x = dviab, y = growth)) +
  geom_hline(yintercept = 1, colour = "grey55", linetype = 2, linewidth = 0.4) +
  geom_vline(xintercept = 0, colour = "grey55", linetype = 2, linewidth = 0.4) +
  geom_point(aes(fill = assay, shape = class, stroke = sig_growth, colour = sig_growth), size = 3.2) +
  geom_text_repel(aes(label = label), size = 2.4, colour = "grey25",
                  min.segment.length = 0.2, segment.colour = "grey70",
                  box.padding = 0.35, max.overlaps = Inf, seed = 1) +
  scale_fill_manual(values = strain_cols, name = "Contamination") +
  scale_shape_manual(values = class_shapes, name = "Inhibitor class") +
  scale_discrete_manual("stroke", values = sig_stroke, guide = "none") +
  scale_colour_manual(values = sig_col, name = "Effect on growth\n(Dunnett's test)") +
  scale_y_log10(breaks = c(0.125, 0.25, 0.5, 1, 2), labels = c(0.125, 0.25, 0.5, 1, 2)) +
  labs(x = "Change in yeast viability relative to No LAB (percentage points)",
       y = "Yeast growth relative to No LAB (X24/X0)") +
  guides(fill   = guide_legend(override.aes = list(shape = 21, size = 3.5, colour = "grey45", stroke = 0.35), order = 1),
         shape  = guide_legend(override.aes = list(fill = "grey85", size = 3, colour = "grey45", stroke = 0.35), order = 2),
         colour = guide_legend(override.aes = list(shape = 21, fill = "white", size = 3,
                                                   stroke = c(1.3, 0.35)), order = 3)) +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "grey85", fill = NA, linewidth = 0.4),
        axis.text = element_text(colour = "black"),
        legend.title = element_text(size = 8.5), legend.text = element_text(size = 8),
        legend.key.height = unit(0.45, "cm"))
 
fig_vg
 
ggsave("Fig_viability_growth.tiff", fig_vg, width = 18, height = 13, units = "cm",
       dpi = 600, device = "tiff", compression = "lzw", bg = "white")
ggsave("Fig_viability_growth.pdf", fig_vg, width = 18, height = 13, units = "cm",
       device = cairo_pdf, bg = "white")
 
```

\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

# Fig4. bacteria growth

```{r}
{r}
## ---- bacteria-setup ----
library(readxl)
library(tidyr)
library(dplyr)
library(multcomp)
library(ggplot2)
 
## ---- bacteria-data ----
# Bacterial growth ratio (N24/N0, CFU/mL) per strain and replicate.
# First row of the sheet = inhibitor-free control.
bac_raw <- read_excel(
  path      = "bubbleplot_bacteria.xlsx",
  sheet     = 1,
  skip      = 2,
  col_names = c("inhibitor",
                "LF_1",   "LF_2",   "LF_3",
                "LP_1",   "LP_2",   "LP_3",
                "LFLP_1", "LFLP_2", "LFLP_3")
) %>%
  filter(!is.na(inhibitor), inhibitor != "") %>%
  mutate(inhibitor = trimws(gsub(" ", " ", inhibitor)),   # remove non-breaking spaces
         across(-inhibitor, as.numeric))
 
inhibitor_levels <- bac_raw$inhibitor          # order of the spreadsheet; control first
control_name     <- inhibitor_levels[1]
assay_levels     <- c("LF", "LP", "LFLP")
 
bac <- bac_raw %>%
  pivot_longer(-inhibitor, names_to = c("assay", "rep"),
               names_sep = "_", values_to = "ratio") %>%
  mutate(inhibitor = factor(inhibitor, levels = inhibitor_levels),
         assay     = factor(assay, levels = assay_levels))
 
## ---- bacteria-normalization ----
# Each value divided by the mean of the inhibitor-free control of the same assay
bac <- bac %>%
  group_by(assay) %>%
  mutate(ref = mean(ratio[inhibitor == control_name]),
         fc  = ratio / ref) %>%
  ungroup()
 
## ---- bacteria-stats ----
# (1) Effect of each inhibitor on each bacterial assay: one-way ANOVA per assay,
#     Dunnett's test of each inhibitor against the inhibitor-free control.
#     NOTE: the control was run in a different experimental block, so these
#     comparisons include day-to-day variation.
# (2) Differences among assays within each inhibitor block (valid within-block):
#     one-way ANOVA + Tukey's HSD.
# All tests on log-transformed growth ratios.
dun_res <- list(); tuk_res <- list(); anova_res <- list()
 
for (a in assay_levels) {
  d   <- filter(bac, assay == a)
  fit <- aov(log(ratio) ~ inhibitor, data = d)
  dun <- summary(glht(fit, linfct = mcp(inhibitor = "Dunnett")))
  dun_res[[a]] <- data.frame(assay      = a,
                             comparison = names(dun$test$coefficients),
                             log_diff   = unname(dun$test$coefficients),
                             p_adj      = unname(dun$test$pvalues))
}
 
for (inh in inhibitor_levels) {
  d   <- filter(bac, inhibitor == inh)
  fit <- aov(log(ratio) ~ assay, data = d)
  anova_res[[inh]] <- data.frame(inhibitor = inh,
                                 F = summary(fit)[[1]][["F value"]][1],
                                 p = summary(fit)[[1]][["Pr(>F)"]][1])
  tuk <- TukeyHSD(fit)$assay
  tuk_res[[inh]] <- data.frame(inhibitor = inh, comparison = rownames(tuk),
                               log_diff = tuk[, "diff"], p_adj = tuk[, "p adj"])
}
 
bac_dunnett <- bind_rows(dun_res)
bac_anova   <- bind_rows(anova_res)
bac_tukey   <- bind_rows(tuk_res); rownames(bac_tukey) <- NULL
 
knitr::kable(bac_dunnett, digits = 3,
             caption = "Dunnett's test: each inhibitor vs inhibitor-free control, per assay (log ratio)")
knitr::kable(bac_anova, digits = 3,
             caption = "One-way ANOVA among assays within each inhibitor block (log ratio)")
knitr::kable(bac_tukey, digits = 3,
             caption = "Tukey's HSD among assays within each inhibitor block (log ratio)")
 
## ---- bacteria-figure ----
sig <- bac_dunnett %>%
  mutate(inhibitor = sub(" - .*$", "", comparison),
         star      = ifelse(p_adj < 0.05, "*", "")) %>%
  dplyr::select(assay, inhibitor, star) %>%
  mutate(inhibitor = factor(inhibitor, levels = inhibitor_levels),
         assay     = factor(assay, levels = assay_levels))
 
summ <- bac %>%
  group_by(inhibitor, assay) %>%
  summarise(mean = mean(fc), sd = sd(fc), .groups = "drop") %>%
  left_join(sig, by = c("inhibitor", "assay")) %>%
  mutate(star       = ifelse(is.na(star), "", star),
         label_mean = paste0(sprintf("%.2f", mean), star),
         label_sd   = paste0("± ", sprintf("%.2f", sd)),
         log2fc     = log2(mean))
 
lim <- max(abs(summ$log2fc), na.rm = TRUE)
summ <- mutate(summ, text_col = ifelse(abs(log2fc) > 0.6 * lim, "white", "grey10"))
fc_breaks <- c(0.125, 0.25, 0.5, 1, 2, 4, 8)
fc_breaks <- fc_breaks[abs(log2(fc_breaks)) <= lim]
 
assay_labels <- c(LF = "LF", LP = "LP", LFLP = "LF + LP")
 
fig4 <- ggplot(summ, aes(x = inhibitor, y = assay)) +
  geom_point(aes(size = mean, fill = log2fc),
             shape = 21, colour = "grey30", stroke = 0.4) +
  geom_text(aes(label = label_mean, colour = text_col),
            size = 2.7, fontface = "bold") +
  geom_text(aes(label = label_sd), colour = "grey35",
            nudge_y = -0.42, size = 2.3) +
  scale_colour_identity() +
  scale_size_continuous(range = c(11, 19), guide = "none") +
  scale_fill_distiller(palette = "RdBu", direction = 1, limits = c(-lim, lim),
                       breaks = log2(fc_breaks), labels = fc_breaks,
                       name = "Bacterial growth\n(relative to control)") +
  scale_y_discrete(limits = rev(assay_levels), labels = assay_labels,
                   expand = expansion(add = 0.6)) +
  scale_x_discrete(guide = guide_axis(angle = 40)) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 10) +
  theme(panel.grid.major = element_line(colour = "grey92", linewidth = 0.3),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "grey85", fill = NA, linewidth = 0.4),
        axis.text   = element_text(colour = "black"),
        legend.title = element_text(size = 8.5),
        plot.margin = margin(12, 8, 4, 4),
        legend.key.height = unit(1.2, "cm"))
 
fig4
 
ggsave("Fig4_bacteria.tiff", fig4, width = 22, height = 9, units = "cm",
       dpi = 600, device = "tiff", compression = "lzw", bg = "white")
ggsave("Fig4_bacteria.pdf", fig4, width = 22, height = 9, units = "cm",
       device = cairo_pdf, bg = "white")
```

\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

# Fig5. metabolites

```{r}
# ---- met-setup ----
library(readxl)
library(tidyr)
library(dplyr)
library(multcomp)
library(ggplot2)
 
## ---- met-data ----
# metabolites.xlsx: one sheet per metabolite (ethanol, glycerol, acetate, lactate),
# rows = inhibitor conditions (control first), columns = assay x replicate.
inh_lab <- c("Control", "HMF", "Furfural", "Acetic acid", "Formic acid",
             "Glycolic acid", "Levulinic acid", "Ferulic acid",
             "Syringaldehyde", "Vanillin")
assay_levels <- c("none", "LF", "LP", "LFLP")
 
read_met <- function(sheet) {
  read_excel("metabolites.xlsx", sheet = sheet, skip = 2,
             col_names = c("inhibitor", paste0(rep(assay_levels, each = 3), "_", 1:3))) %>%
    filter(!is.na(inhibitor), inhibitor != "") %>%
    mutate(inhibitor = inh_lab, across(-inhibitor, as.numeric)) %>%
    pivot_longer(-inhibitor, names_to = c("assay", "rep"), names_sep = "_",
                 values_to = "conc") %>%
    mutate(metabolite = sheet)
}
 
met <- bind_rows(lapply(c("ethanol", "glycerol", "acetate", "lactate"), read_met)) %>%
  filter(!is.na(conc)) %>%
  mutate(inhibitor  = factor(inhibitor, levels = inh_lab),
         assay      = factor(assay, levels = assay_levels),
         metabolite = factor(metabolite, levels = c("ethanol", "glycerol", "acetate", "lactate")))
 
## ---- met-stats ----
# Within each inhibitor block and metabolite:
#  - ethanol, glycerol, acetate: one-way ANOVA + Dunnett vs No LAB
#  - lactate (not produced without bacteria): Tukey among LF, LP and LF + LP
dun_res <- list(); tuk_res <- list()
for (m in levels(met$metabolite)) for (inh in inh_lab) {
  d <- droplevels(filter(met, metabolite == m, inhibitor == inh))
  if (nrow(d) == 0) next
  if ("none" %in% d$assay) {
    dun <- summary(glht(aov(conc ~ assay, data = d), linfct = mcp(assay = "Dunnett")))
    dun_res[[paste(m, inh)]] <- data.frame(metabolite = m, inhibitor = inh,
                                           comparison = names(dun$test$coefficients),
                                           diff  = unname(dun$test$coefficients),
                                           p_adj = unname(dun$test$pvalues))
  }
  dc  <- droplevels(filter(d, assay != "none"))
  tuk <- TukeyHSD(aov(conc ~ assay, data = dc))$assay
  tuk_res[[paste(m, inh)]] <- data.frame(metabolite = m, inhibitor = inh,
                                         comparison = rownames(tuk),
                                         diff = tuk[, "diff"], p_adj = tuk[, "p adj"])
}
met_dunnett <- bind_rows(dun_res); rownames(met_dunnett) <- NULL
met_tukey   <- bind_rows(tuk_res); rownames(met_tukey)   <- NULL
 
knitr::kable(met_dunnett, digits = 3, caption = "Dunnett's test vs No LAB (g/L)")
knitr::kable(met_tukey,   digits = 3, caption = "Tukey's HSD among contaminated assays (g/L)")
 
## ---- met-figure ----
# Style of the original figure: black symbols per assay, coloured strip with the
# metabolite name on the right of each panel, light coloured panel border.
library(patchwork)
 
assay_labels <- c(none = "no LAB", LF = "LF", LP = "LP", LFLP = "LF + LP")
assay_shapes <- c(none = 16, LF = 17, LP = 15, LFLP = 18)
met_cols <- c(ethanol = "#41638E", glycerol = "#3C765B",
              acetate = "#A16A2D", lactate  = "#8E3C3C")
x_labels <- c("control", "HMF (1g.L-1)", "Furfural (3g.L-1)", "Acetic acid (4g.L-1)",
              "Formic acid (0.75g.L-1)", "Glycolic acid (0.75g.L-1)",
              "Levulinic acid (3g.L-1)", "Ferulic acid (0.75g.L-1)",
              "Syringaldehyde (0.75g.L-1)", "Vanillin (0.75g.L-1)")
x_expr <- lapply(x_labels, function(l) {
  if (!grepl("g.L-1", l)) return(l)
  parse(text = paste0('"', sub("g.L-1\\)", "", l), 'g.L"^-1*")"'))[[1]]
})
 
sig <- met_dunnett %>%
  mutate(assay = sub(" - none", "", comparison),
         star  = ifelse(p_adj < 0.05, "*", "")) %>%
  dplyr::select(metabolite, inhibitor, assay, star)
 
summ <- met %>%
  group_by(metabolite, inhibitor, assay) %>%
  summarise(mean = mean(conc), sd = sd(conc), .groups = "drop") %>%
  left_join(mutate(sig, metabolite = factor(metabolite, levels = levels(met$metabolite)),
                   inhibitor = factor(inhibitor, levels = inh_lab),
                   assay = factor(assay, levels = assay_levels)),
            by = c("metabolite", "inhibitor", "assay")) %>%
  mutate(star = ifelse(is.na(star), "", star))
 
dodge <- position_dodge(width = 0.65)
 
met_panel <- function(m, show_x = FALSE, show_legend = FALSE) {
  d <- filter(summ, metabolite == m) %>%
    mutate(strip = paste0(toupper(m), "\n(g.L\u207b\u00b9)"))
  ggplot(d, aes(x = inhibitor, y = mean, shape = assay, group = assay)) +
    geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.15,
                  linewidth = 0.35, position = dodge) +
    geom_point(aes(size = assay), position = dodge) +
    scale_size_manual(values = c(none = 2.6, LF = 2.6, LP = 2.4, LFLP = 3.3), guide = "none") +
    geom_text(aes(y = mean + sd, label = star), size = 3.3, vjust = -0.15,
              position = dodge, show.legend = FALSE) +
    scale_shape_manual(values = assay_shapes, labels = assay_labels, name = NULL, drop = FALSE) +
    scale_x_discrete(labels = x_expr, drop = FALSE) +
    scale_y_continuous(expand = expansion(mult = c(0.06, 0.16))) +
    guides(shape = guide_legend(override.aes = list(size = c(3.6, 3.6, 3.3, 4.4)))) +
    facet_grid(strip ~ .) +
    labs(x = NULL, y = NULL) +
    theme_minimal(base_size = 10) +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major = element_line(colour = "grey92", linewidth = 0.3),
          panel.border = element_rect(colour = scales::alpha(met_cols[[m]], 0.45),
                                      fill = NA, linewidth = 1),
          strip.background = element_rect(fill = met_cols[[m]], colour = NA),
          strip.text.y = element_text(colour = "white", face = "bold", size = 11,
                                      angle = 90, lineheight = 0.9),
          axis.text = element_text(colour = "black"),
          axis.text.x = if (show_x) element_text(angle = 40, hjust = 1) else element_blank(),
          legend.position = if (show_legend) "top" else "none",
          legend.text = element_text(size = 11),
          plot.margin = margin(1, 4, 1, 4))
}
 
fig5 <- met_panel("ethanol", show_legend = TRUE) /
  met_panel("glycerol") /
  met_panel("acetate") /
  met_panel("lactate", show_x = TRUE)
 
fig5
 
ggsave("Fig5_metabolites.tiff", fig5, width = 17, height = 18, units = "cm",
       dpi = 600, device = "tiff", compression = "lzw", bg = "white")
ggsave("Fig5_metabolites.pdf", fig5, width = 17, height = 18, units = "cm",
       device = cairo_pdf, bg = "white")
 
```

```{r}
## ---- met-ethanol-lactate ----
# Ethanol loss (No LAB minus contaminated) vs lactate produced, per block and assay
eth_lac <- met %>%
  group_by(metabolite, inhibitor, assay) %>%
  summarise(v = mean(conc), .groups = "drop") %>%
  pivot_wider(names_from = metabolite, values_from = v) %>%
  group_by(inhibitor) %>%
  mutate(ethanol_loss = ethanol[assay == "none"] - ethanol) %>%
  ungroup() %>%
  filter(assay != "none") %>%
  mutate(expected_loss = 0.511 * lactate)   # ~0.51 g ethanol forgone per g lactate

knitr::kable(eth_lac %>% dplyr::select(inhibitor, assay, ethanol_loss, lactate, expected_loss),
             digits = 2, caption = "Observed ethanol loss vs loss expected from lactate")

cor.test(eth_lac$ethanol_loss, eth_lac$lactate)                        # all conditions
with(filter(eth_lac, inhibitor != "Formic acid"),
     cor.test(ethanol_loss, lactate))                                   # excluding formic acid
summary(lm(ethanol_loss ~ lactate, data = eth_lac))                     # slope vs 0.511
```

```{r}
## ---- eth-lac-setup ----
# Run AFTER the metabolites chunks (uses the object 'met').
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)

## ---- eth-lac-data ----
# Ethanol loss = mean ethanol (No LAB) - mean ethanol (contaminated), same block.
# Expected loss if the glucose converted to lactate had been fermented by yeast:
# ~0.511 g ethanol per g lactate (same net value for homo- and heterofermentative LAB).
eth_lac <- met %>%
  group_by(metabolite, inhibitor, assay) %>%
  summarise(v = mean(conc), .groups = "drop") %>%
  pivot_wider(names_from = metabolite, values_from = v) %>%
  group_by(inhibitor) %>%
  mutate(ethanol_loss = ethanol[assay == "none"] - ethanol) %>%
  ungroup() %>%
  filter(assay != "none") %>%
  mutate(expected_loss = 0.511 * lactate,
         formic = inhibitor == "Formic acid",
         assay  = factor(recode(as.character(assay), LFLP = "LF + LP"),
                         levels = c("LF", "LP", "LF + LP")))

knitr::kable(eth_lac %>% dplyr::select(inhibitor, assay, ethanol_loss, lactate, expected_loss),
             digits = 2, caption = "Observed ethanol loss vs loss expected from lactate")

## ---- eth-lac-stats ----
ct_all <- cor.test(eth_lac$ethanol_loss, eth_lac$lactate)
ct_nof <- with(filter(eth_lac, !formic), cor.test(ethanol_loss, lactate))
fit    <- lm(ethanol_loss ~ lactate, data = filter(eth_lac, !formic))
ct_all; ct_nof; summary(fit)

## ---- eth-lac-figure ----
strain_cols <- c("LF" = "#F9BC3C", "LP" = "#F8866D", "LF + LP" = "#7A5340")
lab_r <- sprintf("r = %.2f (all conditions)\nr = %.2f (excluding formic acid)",
                 ct_all$estimate, ct_nof$estimate)

figS1 <- ggplot(eth_lac, aes(x = lactate, y = ethanol_loss)) +
  geom_hline(yintercept = 0, colour = "grey70", linewidth = 0.3) +
  geom_abline(intercept = 0, slope = 0.511, linetype = 2, colour = "grey35") +
  geom_smooth(data = filter(eth_lac, !formic), method = "lm", se = TRUE,
              colour = "black", fill = "grey85", linewidth = 0.5, formula = y ~ x) +
  geom_point(aes(fill = assay, shape = formic), size = 3, colour = "grey25", stroke = 0.4) +
  geom_text_repel(aes(label = inhibitor), size = 2.3, colour = "grey30",
                  segment.colour = "grey75", min.segment.length = 0.2,
                  max.overlaps = Inf, seed = 1) +
  annotate("text", x = 0.85, y = 1.75, label = lab_r, hjust = 0, size = 3) +
  annotate("text", x = 2.65, y = 0.511 * 2.65, label = "expected\n(0.51 g/g)",
           hjust = 0, vjust = 1.2, size = 2.6, colour = "grey35") +
  scale_fill_manual(values = strain_cols, name = "Contamination") +
  scale_shape_manual(values = c(`FALSE` = 21, `TRUE` = 24),
                     labels = c(`FALSE` = "Other inhibitors", `TRUE` = "Formic acid"),
                     name = NULL) +
  guides(fill = guide_legend(override.aes = list(shape = 21, size = 3.5))) +
  scale_x_continuous(limits = c(0.8, 2.9)) +
  labs(x = expression("Lactate produced (g" %.% "L"^-1 * ")"),
       y = expression("Ethanol loss relative to No LAB (g" %.% "L"^-1 * ")")) +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(),
        panel.border = element_rect(colour = "grey85", fill = NA, linewidth = 0.4),
        axis.text = element_text(colour = "black"))

figS1

ggsave("FigS1_ethanol_lactate.tiff", figS1, width = 17, height = 12, units = "cm",
       dpi = 600, device = "tiff", compression = "lzw", bg = "white")
ggsave("FigS1_ethanol_lactate.pdf", figS1, width = 17, height = 12, units = "cm",
       device = cairo_pdf, bg = "white")
```

\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

# Fig6. pH

```{r}
# ---- ph-setup ----
library(tidyr)
library(dplyr)
library(multcomp)
library(ggplot2)
 
## ---- ph-data ----
# ph.csv: final pH (24 h), one row per inhibitor (Table 1 order, control first),
# columns = assay x replicate (none_1 ... LFLP_3).
inh_lab <- c("Control", "HMF", "Furfural", "Acetic acid", "Formic acid",
             "Glycolic acid", "Levulinic acid", "Ferulic acid",
             "Syringaldehyde", "Vanillin")
assay_levels <- c("none", "LF", "LP", "LFLP")
 
ph <- read.csv("ph.csv", check.names = FALSE) %>%
  mutate(inhibitor = inh_lab) %>%
  pivot_longer(-inhibitor, names_to = c("assay", "rep"), names_sep = "_", values_to = "pH") %>%
  mutate(inhibitor = factor(inhibitor, levels = inh_lab),
         assay     = factor(assay, levels = assay_levels))
 
## ---- ph-stats ----
# Within each inhibitor block: Dunnett vs No LAB; Tukey among contaminated assays.
dun_res <- list(); tuk_res <- list()
for (inh in inh_lab) {
  d   <- filter(ph, inhibitor == inh)
  dun <- summary(glht(aov(pH ~ assay, data = d), linfct = mcp(assay = "Dunnett")))
  dun_res[[inh]] <- data.frame(inhibitor = inh, comparison = names(dun$test$coefficients),
                               diff = unname(dun$test$coefficients),
                               p_adj = unname(dun$test$pvalues))
  tuk <- TukeyHSD(aov(pH ~ assay, data = droplevels(filter(d, assay != "none"))))$assay
  tuk_res[[inh]] <- data.frame(inhibitor = inh, comparison = rownames(tuk),
                               diff = tuk[, "diff"], p_adj = tuk[, "p adj"])
}
ph_dunnett <- bind_rows(dun_res); rownames(ph_dunnett) <- NULL
ph_tukey   <- bind_rows(tuk_res); rownames(ph_tukey)   <- NULL
 
knitr::kable(ph_dunnett, digits = 3, caption = "Final pH: Dunnett's test vs No LAB")
knitr::kable(ph_tukey,   digits = 3, caption = "Final pH: Tukey's HSD among contaminated assays")
 
## ---- ph-figure ----
sig <- ph_dunnett %>%
  mutate(assay = sub(" - none", "", comparison),
         star  = ifelse(p_adj < 0.05, "*", "")) %>%
  dplyr::select(inhibitor, assay, star) %>%
  mutate(inhibitor = factor(inhibitor, levels = inh_lab),
         assay     = factor(assay, levels = assay_levels))
 
summ <- ph %>%
  group_by(inhibitor, assay) %>%
  summarise(mean = mean(pH), sd = sd(pH), .groups = "drop") %>%
  left_join(sig, by = c("inhibitor", "assay")) %>%
  mutate(star       = ifelse(is.na(star), "", star),
         label_mean = paste0(sprintf("%.2f", mean), star),
         label_sd   = paste0("± ", sprintf("%.2f", sd)),
         text_col   = ifelse(mean < 4.3 | mean > 5.6, "white", "grey10"))
 
assay_labels <- c(none = "No LAB", LF = "LF", LP = "LP", LFLP = "LF + LP")
 
fig6 <- ggplot(summ, aes(x = inhibitor, y = assay, fill = mean)) +
  geom_tile(colour = "white", linewidth = 1.2) +
  geom_text(aes(label = label_mean, colour = text_col),
            nudge_y = 0.13, size = 3.2, fontface = "bold") +
  geom_text(aes(label = label_sd, colour = text_col),
            nudge_y = -0.20, size = 2.4) +
  scale_colour_identity() +
  scale_fill_distiller(palette = "RdBu", direction = 1, limits = c(3.6, 6.2),
                       name = "Final pH") +
  scale_y_discrete(limits = rev(assay_levels), labels = assay_labels) +
  scale_x_discrete(guide = guide_axis(angle = 40)) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 10) +
  theme(panel.grid  = element_blank(),
        axis.text   = element_text(colour = "black"),
        legend.key.height = unit(1.2, "cm"))
 
fig6
 
ggsave("Fig6_pH.tiff", fig6, width = 22, height = 9, units = "cm",
       dpi = 600, device = "tiff", compression = "lzw", bg = "white")
ggsave("Fig6_pH.pdf", fig6, width = 22, height = 9, units = "cm",
       device = cairo_pdf, bg = "white")
 
## ---- ph-lactate ----
# Relationship between pH drop and lactate produced (contaminated cultures)
# Requires the 'met' object from the metabolites chunks.
if (exists("met")) {
  lac <- met %>% filter(metabolite == "lactate") %>%
    group_by(inhibitor, assay) %>% summarise(lactate = mean(conc), .groups = "drop")
  dph <- summ %>% group_by(inhibitor) %>%
    mutate(delta_pH = mean - mean[assay == "none"]) %>% ungroup() %>%
    filter(assay != "none") %>%
    left_join(lac, by = c("inhibitor", "assay"))
  print(cor.test(dph$delta_pH, dph$lactate))
}
 
```
